\part{Morphology}

\chapter{Nouns}

\section{Formation}

\latscript{-CI} denotes those exercising a profession or habitual involvement with the object or concept expressed by the underived noun.

\begin{lingexample}
	\ex \latscript{balıqçı} `fisherman' \derivedfrom{} \latscript{balıq} `fish' \\
		\latscript{oquwcu} `reader' \derivedfrom{} \latscript{oquw} `reading' \\
		\latscript{temirci} `(black)smith' \derivedfrom{} \latscript{temir} `iron'
\end{lingexample}

\latscript{-DAş} denotes entities with a common attachment to the attribute, item or concept expressed by the underived noun.

\begin{lingexample}
	\ex \latscript{qarındaş} `sibling' \derivedfrom{} \latscript{qarın} `womb' \\
		\latscript{wätändeş} `fellow countryman' \derivedfrom{} \latscript{wätän} `homeland' \\
		\latscript{yoldaş} `fellow traveller, companion, comrade' \derivedfrom{} \latscript{yol} `road, path'
\end{lingexample}

\section{Inflection}

Nouns are inflected for number, possession and case.
When multiple such suffixes co-occur, they are added in the order of number, possession, then case.

\begin{lingexample}
	\ex \gll üy-ler-imiz-den \\
		house-\Pl{}-\Poss{}.\Fpl{}-\Abl{} \\
		\glt `from our houses'
\end{lingexample}

\subsection{Number}

Singular nouns are unmarked, while plural nouns are formed by suffixing \latscript{-lAr} onto the singular stem.
Generally, only count nouns are pluralised.

\begin{lingexample}
	\ex \latscript{almalar} `apples' \derivedfrom{} \latscript{alma} `apple' \\
		\latscript{balalar} `children' \derivedfrom{} \latscript{bala} `child' \\
		\latscript{üyler} `houses' \derivedfrom{} \latscript{üy} `house'
\end{lingexample}

However, mass nouns can also be pluralised, with the plural suffix denoting non-specificity or discrete masses of the referent, rather than individual entities.

\begin{lingexample}
	\ex \latscript{ballar} `honeys, (jars/types/etc.\ of) honey' \derivedfrom{} \latscript{bal} `honey' \\
		\latscript{suwlar} `waters, (bodies/masses/etc.\ of) water' \derivedfrom{} \latscript{suw} `water'
\end{lingexample}

The plural suffix is never used if plurality is indicated by other means, such as by numerals or adverbs of quantity.

\begin{lingexample}
	\ex \latscript{beş oquwcu} `five students' \\
		\latscript{köp kişi} `many people'
\end{lingexample}

\subsection{Possession}

Possession is expressed by the suffixes in table~\ref{tab:possessive-suffixes}, which encode the person and number of the possessor.
The third person is the exception, which does not distinguish between a singular or plural possessor.

\begin{smalltable}[\label{tab:possessive-suffixes}]{Possessive suffixes}{c|c|c|c}
	& \textbf{\First{}} & \textbf{\Second{}} & \textbf{\Third{}} \\
	\hline
	\textbf{\Sg{}} & \latscript{-(I)m} & \latscript{-(I)ñ} & \multirow{2}{*}{\latscript{-(s)I}} \\
	\cline{1-3}
	\textbf{\Pl{}} & \latscript{-(I)mIz} & \latscript{-(I)ñIz} & \\
\end{smalltable}

If present, the possessor would be declined in the genitive case.

\begin{lingexample}
	\ex \gll seniñ ayağ-ıñ \\
		\Ssg{}.\Gen{} foot-\Poss{}.\Ssg{} \\
		\glt `your foot'
\end{lingexample}

Because they act as determiners, the possessive suffixes confer definiteness onto the nominals to which they are attached.
Consequently, if said nominal is used as a direct object, it must be declined in the accusative case.

\begin{lingexample}
	\ex \gll Öz qalqan-ı-n tut-ma-dı. \\
		\Refl{} shield-\Poss{}.\Third{}-\Acc{} hold-\Neg{}-\Pst{}.\Def{} \\
		\glt `He wasn\apostrophe{}t holding his (own) shield.'
\end{lingexample}

\subsection{Case}

There are six cases in Interturkic, whose suffixes are given in table~\ref{tab:case-suffixes}.
The accusative\apostrophe{}s vowel is optionally dropped after the third-person possessive suffix.
The dative, locative, and ablative also have different forms when placed after a possessive suffix.

\begin{smalltable}[\label{tab:case-suffixes}]{Case suffixes}{c|c|c|c|c|c|c|c}
	\multicolumn{2}{|c|}{} & \textbf{\Nom{}} & \textbf{\Acc{}} & \textbf{\Gen{}} & \textbf{\Dat{}} & \textbf{\Loc{}} & \textbf{\Abl{}} \\
	\hline
	\multicolumn{2}{|c|}{\textbf{Normal}} & \multirow{3}{*}{\latscript{-\zero{}}} & \multirow{2}{*}{\latscript{-nI}} & \multirow{3}{*}{\latscript{-nIñ}} & \latscript{-GA} & \latscript{-DA} & \latscript{-DAn} \\
	\cline{1-2} \cline{6-8}
	\multirow{2}{*}{\textbf{Possessive}} & \textbf{\First{}/\Second{}} & & & & \multirow{2}{*}{\latscript{-(n)A}} & \multirow{2}{*}{\latscript{-(n)dA}} & \multirow{2}{*}{\latscript{-(n)dAn}} \\
	\cline{2-2} \cline{4-4}
	& \textbf{\Third{}} & & \latscript{-n[I]} & & & & \\
\end{smalltable}

The nominative case marks the subject or predicative complement of a sentence (see example~\ref{ex:nom-subject-predicative}), as well as indefinite direct objects (see example~\ref{ex:nom-indef-direct-obj}).
Otherwise, the accusative case must be used.

\begin{lingexample}
	\ex \label{ex:nom-subject-predicative} \gll Ol yaqşı yoldaş. \\
		\Tsg{} good friend-\Sg{} \\
		\glt `He\apostrophe{}s a good friend.'
	\ex \label{ex:nom-indef-direct-obj} \gll Men taw-lar kör-dü-m. \\
		\Fsg{} mountain-\Pl{} see-\Pst{}.\Def{}-\Fsg{} \\
		\glt `I saw mountains.'
\end{lingexample}

The accusative case marks direct objects if and onlt if they are considered definite (compare examples-\ref{ex:acc-def-direct-obj} and \ref{ex:no-acc-indef-dir-obj}).

\begin{lingexample}
	\ex \label{ex:acc-def-direct-obj} \gll Sen at-nı kör-dü-ñ mü? \\
		\Ssg{} horse-\Acc{} see-\Pst{}.\Def{}-\Ssg{} \Q{} \\
		\glt `Did you see \emph{the} horse?'
	\ex \label{ex:no-acc-indef-dir-obj} \gll Sen at kör-dü-ñ mü? \\
		\Ssg{} horse see-\Pst{}.\Def{}-\Ssg{} \Q{} \\
		\glt `Did you see \emph{a} horse?'
\end{lingexample}

The genitive case expresses a possessive or qualifying relationship of the marked nominal over another.
The corresponding possessive suffix is then suffixed onto the qualified nominal.

\begin{lingexample}
	\ex \gll Olar-nıñ üy-ler-i aq. \\
		\Tpl{}-\Gen{} house-\Pl{}-\Poss{}.\Third{} white \\
		\glt `Their houses are white.'
\end{lingexample}

The dative case marks indirect objects (see example~\ref{ex:dat-indirect-object}), the complements of some verbs (see example~\ref{ex:dat-verb-complement}), and the point towards which an action is directed (see example~\ref{ex:dat-destination}).

\begin{lingexample}
	\ex \label{ex:dat-indirect-object} \gll Arslan maña üç balıq ber-di. \\
		Arslan \Fsg{}.\Dat{} three fish give-\Pst{}.\Def{} \\
		\glt `Arslan gave me three fish.'
	\ex \label{ex:dat-verb-complement} \gll Siz ol xanlıq-qa kömek ber-me-yecek-siz mi? \\
		\Spl{} \Dist{} khanate-\Dat{} help give-\Neg{}-\Fut{}.\Def{}-\Spl{} \Q{} \\
		\glt `Aren\apostrophe{}t you going to help that khanate?'
	\ex \label{ex:dat-destination} \gll Biz Azärbaycan-ğa bar-dı-q. \\
		\Fpl{} Azerbaijan-\Dat{} go-\Pst{}.\Def{}-\Fpl{} \\
		\glt `We went to Azerbaijan.'
\end{lingexample}

The locative case expresses location at a physical location (see example~\ref{ex:loc-location}), within an abstract concept (see example~\ref{ex:loc-abstract}), or at a point in time (see example~\ref{ex:loc-time}).

\begin{lingexample}
	\ex \label{ex:loc-location} \gll Üy-ümüz-de tuz yoq. \\
		house-\Poss{}.\Fpl{}-\Loc{} salt absent \\
		\glt `There\apostrophe{}s no salt in my house.'
	\ex \label{ex:loc-abstract} \gll Eşğal-da-mız mı? \\
		occupation-\Loc{}-\Cop{}.\Prs{}.\Fpl{} \Q{} \\
		\glt `Are we under occupation?'
	\ex \label{ex:loc-time} \gll Taw-dan sä\apostrophe{}at altı-da çıq-tı-lar. \\
		mountain-\Abl{} hour six-\Loc{} go\_out-\Pst{}.\Def{}-\Tpl{} \\
		\glt `They left the mountain at six o\apostrophe{}clock'
\end{lingexample}

\chapter{Adjectives}

\section{Formation}

\latscript{-GI} forms relational adjectives from uninflected nominals (see example~\ref{ex:gi-relational}).
This suffix can also attach to nominals in the genitive (as \latscript{-KI}, see example~\ref{ex:gi-genitive}) or locative (see example~\ref{ex:gi-locative}), forming adjectives pertaining to possession or location.

\begin{exe}
	\ex \label{ex:gi-relational} \latscript{bugüngü} `today\apostrophe{}s, present-day' \derivedfrom{} \latscript{bugün} `today' \\
		\latscript{qışqı} `wintry, hibernal' \derivedfrom{} \latscript{qış} `winter'
	\ex \label{ex:gi-genitive} \latscript{meniñki} `mine' \derivedfrom{} \latscript{meniñ} `my' \\
		\latscript{quşnuñqu} `that which is a/the bird\apostrophe{}s' \derivedfrom{} \latscript{quşnuñ} `bird\apostrophe{}s'
	\ex \label{ex:gi-locative} \latscript{Astanadağı} `that which is in Astana' \derivedfrom{} \latscript{Astanada} `in Astana' \\
		\latscript{üyümdegi} `that which is in my house' \derivedfrom{} \latscript{üyümde} `in my house'
\end{exe}

If adjectives in the latter two groups are inflected for case without any other intervening morphology, the possessive set of case markers must be used.

\begin{exe}
	\ex \latscript{bizniñkinden} `from ours' \derivedfrom{} \latscript{bizniñki} `ours'
\end{exe}

Otherwise, adjectives formed from \latscript{-GI} can be used attributively.

\begin{lingexample}
	\ex \gll Bu awıl-da-ğı ağaç-lar uzun. \\
		\Prox{} village-\Loc{}-\Rel{} tree-\Pl{} tall \\
		`The trees in this village are tall.'
\end{lingexample}

\section{Comparison}

\chapter{Numerals}

\section{Cardinal numerals}

\section{Collective numerals}

\section{Ordinal numerals}

\section{Distributive numerals}

\chapter{Pronouns}

\section{Personal pronouns}

\begin{smalltable}{Declension of personal pronouns}{c|c|c|c|c}
	& \textbf{\Fsg{}} & \textbf{\Ssg{}} & \textbf{\Fpl{}} & \textbf{\Spl{}} \\
	\hline
	\textbf{\Nom{}} & \latscript{men} & \latscript{sen} & \latscript{biz} & \latscript{siz} \\
	\hline
	\textbf{\Acc{}} & \latscript{meni} & \latscript{seni} & \latscript{bizni} & \latscript{sizni} \\
	\hline
	\textbf{\Gen{}} & \latscript{meniñ} & \latscript{seniñ} & \latscript{bizniñ} & \latscript{sizniñ} \\
	\hline
	\textbf{\Dat{}} & \latscript{maña} & \latscript{saña} & \latscript{bizge} & \latscript{sizge} \\
	\hline
	\textbf{\Loc{}} & \latscript{mende} & \latscript{sende} & \latscript{bizde} & \latscript{sizde} \\
	\hline
	\textbf{\Abl{}} & \latscript{menden} & \latscript{senden} & \latscript{bizden} & \latscript{sizden} \\
\end{smalltable}

\section{Reflexive pronouns}

\begin{smalltable}{Declension of reflexive pronouns}{c|c|c|c|c|c|c}
	& \textbf{\Fsg{}} & \textbf{\Ssg{}} & \textbf{\Tsg{}} & \textbf{\Fpl{}} & \textbf{\Spl{}} & \textbf{\Tpl{}} \\
	\hline
	\textbf{\Nom{}} & \latscript{özüm} & \latscript{özüñ} & \latscript{özü} & \latscript{özümüz} & \latscript{özüñüz} & \latscript{özleri} \\
	\hline
	\textbf{\Acc{}} & \latscript{özümü} & \latscript{özüñü} & \latscript{özün[ü]} & \latscript{özümüznü} & \latscript{özüñüznü} & \latscript{özlerin[i]} \\
	\hline
	\textbf{\Gen{}} & \latscript{özümüñ} & \latscript{özüñüñ} & \latscript{özünüñ} & \latscript{özümüznüñ} & \latscript{özüñüznüñ} & \latscript{özleriniñ} \\
	\hline
	\textbf{\Dat{}} & \latscript{özüme} & \latscript{özüñe} & \latscript{özüne} & \latscript{özümüze} & \latscript{özüñüze} & \latscript{özlerine} \\
	\hline
	\textbf{\Loc{}} & \latscript{özümde} & \latscript{özüñde} & \latscript{özünde} & \latscript{özümüzde} & \latscript{özüñüzde} & \latscript{özlerinde} \\
	\hline
	\textbf{\Abl{}} & \latscript{özümden} & \latscript{özüñden} & \latscript{özünden} & \latscript{özümüzden} & \latscript{özüñüzden} & \latscript{özlerinden} \\
\end{smalltable}

\section{Demonstrative pronouns}

The demonstrative pronouns are derived from the determiners \latscript{bu} `this (proximal)', \latscript{şu} `this/that (medial)', and \latscript{ol} `that (distal)'.

The proximal \latscript{bu} refers to objects in the immediate vincinity of the speaker and listeners, oftentimes being the general topic of conversation (see example~\ref{ex:proximal-determiner}).
The medial \latscript{şu} is used to draw attention to something for the first time, often accompanied by a gesture like pointing (see example~\ref{ex:medial-determiner}).
The distal \latscript{ol} refers to objects outside of the immediate vicinity of the speaker and listener that been previously mentioned (see example~\ref{ex:distal-determiner}).

\begin{lingexample}
	\ex \label{ex:proximal-determiner} \gll Bu tawliq biz-den uzaq emes. \\
		\Prox{} upland \Fpl{}-\Abl{} far \Cop{}.\Neg{} \\
		\glt `This upland (that I just mentioned) isn\apostrophe{}t far from us.'
	\ex \label{ex:medial-determiner} \gll Şu üy-ler biz-niñ-ki. \\
		\Med{} house-\Pl{} \Fpl{}-\Gen{}-\Rel{} \\
		\glt `These/those houses (to which I\apostrophe{}m pointing) are ours.'
	\ex \label{ex:distal-determiner} \gll Ol çöl-de qum yoq. \\
		\Dist{} desert-\Loc{} sand absent. \\
		\glt `There\apostrophe{}s no sand in that desert (which is the general topic).'
\end{lingexample}

When used attributively, these determiners can modify nominal phrases in similar fashion as adjectives (see example~\ref{ex:determiner-as-attribute}).
In this role, the determiners remain in their bare forms, with the modified nominal phrases bearing any and all inflection.
When used as pronouns, these determiners are inflected for number and case (see example~\ref{ex:determiner-as-pronoun}).

\begin{lingexample}
	\ex \label{ex:determiner-as-attribute} \gll Sen bu kino-nu kör-dü-ñ mü? \\
		\Tsg{} \Prox{} movie-\Acc{} see-\Pst{}.\Def{}-\Tsg{} \Q{} \\
		\glt `Have you seen this movie?'
	\ex \label{ex:determiner-as-pronoun} \gll Ondan qorq-ma-y-man. \\
		\Dist{}.\Abl{} be\_afraid-\Neg{}-\Prs{}-\Fsg{} \\
		\glt `I\apostrophe{}m not scared of that.'
\end{lingexample}

\begin{smalltable}{Declension of demonstrative pronouns}{c|c|c|c|c|c|c}
	& \textbf{\Prox{}.\Sg{}} & \textbf{\Med{}.\Sg{}} & \textbf{\Dist{}.\Sg{}} & \textbf{\Prox{}.\Pl{}} & \textbf{\Med{}.\Pl{}} & \textbf{\Dist{}.\Pl{}} \\
	\hline
	\textbf{\Nom{}} & \latscript{bu} & \latscript{şu} & \latscript{ol} & \latscript{bular} & \latscript{şular} & \latscript{olar} \\
	\hline
	\textbf{\Acc{}} & \latscript{bunu} & \latscript{şunu} & \latscript{onu} & \latscript{bularnı} & \latscript{şularnı} & \latscript{olarnı} \\
	\hline
	\textbf{\Gen{}} & \latscript{bunuñ} & \latscript{şunuñ} & \latscript{onuñ} & \latscript{bularnıñ} & \latscript{şularnıñ} & \latscript{olarnıñ} \\
	\hline
	\textbf{\Dat{}} & \latscript{buña} & \latscript{şuña} & \latscript{oña} & \latscript{bularğa} & \latscript{şularğa} & \latscript{olarğa} \\
	\hline
	\textbf{\Loc{}} & \latscript{bunda} & \latscript{şunda} & \latscript{onda} & \latscript{bularda} & \latscript{şularda} & \latscript{olarda} \\
	\hline
	\textbf{\Abl{}} & \latscript{bundan} & \latscript{şundan} & \latscript{ondan} & \latscript{bulardan} & \latscript{şulardan} & \latscript{olardan} \\
\end{smalltable}

\chapter{Postpositions}

\chapter{Verbs}

\section{Formation}

\section{Stems}

Every verb has two stems: positive and negative.
Positive stems express the basic idea of the action and are formed by simply removing the infinitive (see example~\ref{ex:positive-stem-formation}).
This is also the verb\apostrophe{}s basic stem.
Negative stems express the non-performance of an action and are formed by suffixing \latscript{-mA} to the positive stem (see example~\ref{ex:negative-stem-formation}).

\begin{lingexample}
	\ex \label{ex:positive-stem-formation} \latscript{kör-} `see' \derivedfrom{} \latscript{kör-mek} `to see' \\
		\latscript{yaz-} `write' \derivedfrom{} \latscript{yaz-maq} `to write'
	\ex \label{ex:negative-stem-formation} \latscript{al-ma-} `not take' \derivedfrom{} \latscript{al-} `take' \\
		\latscript{ber-me-} `not give' \derivedfrom{} \latscript{ber-} `give'
\end{lingexample}

\section{Non-finite forms}

\subsection{Participles}

\section{Person agreement}

Finite verbs are marked with personal suffixes that encode the subject\apostrophe{}s person and number.
These markers can be of either pronominal or possessive origin (see table~\ref{tab:personal-suffixes}), named as such for their resemblance to the personal pronouns and to the possessive suffixes, respectively.
The choice of which set to use depends on the particular verb paradigm in question, though a handful of paradigms have their own separate set of markers.

\begin{smalltable}[\label{tab:personal-suffixes}]{Personal suffixes}{c|c|c|c|c|c|c}
	& \multicolumn{3}{c|}{\textbf{\Sg{}}} & \multicolumn{3}{c|}{\textbf{\Pl{}}} \\
	\cline{2-7}
	& \textbf{\First{}} & \textbf{\Second{}} & \textbf{\Third{}} & \textbf{\First{}} & \textbf{\Second{}} & \textbf{\Third{}} \\
	\hline
	\textbf{Pronominal} & \latscript{-mAn} & \latscript{-sAn} & \multirow{2}{*}{\latscript{-\zero{}}} & \latscript{-mIz} & \latscript{-sIz} & \multirow{2}{*}{\latscript{[-lAr]}} \\
	\cline{1-3} \cline{5-6}
	\textbf{Possessive} & \latscript{-m} & \latscript{-ñ} & & \latscript{-K} & \latscript{-ñIz} & \\
\end{smalltable}

\section{Finite forms}

\subsection{Indicative}

\subsection{Conditional}

\chapter{Adverbs}
