\part{Phonology}

\chapter{Phonemic inventory}

\section{Vowels}

The Interturkic vowel system is characterised by oppositions in backness, rounding, and height.
These oppositions are vitally important to the process of vowel harmony, and the arrangement of Interturkic\apostrophe{}s nine vowel phonemes along said oppositions in shown below.

\begin{smalltable}{Vowel phonemes}{c|c|c|c|c}
	& \multicolumn{2}{c|}{\textbf{Front}} & \multicolumn{2}{c|}{\textbf{Back}} \\
	\cline{2-5}
	& \textbf{unrounded} & \textbf{rounded} & \textbf{unrounded} & \textbf{rounded} \\
	\hline
	\textbf{High} & i & y & ɯ & u \\
	\hline
	\textbf{Low} & ɛ æ & œ & ɑ & ɔ \\
\end{smalltable}

Dipthongs are not considered to be separate phonemes.
Sequences perceived as rising dipthongs are analysed as /j/ or /w/ followed by a vowel (see example~\ref{ex:rising-diphthongs}).
Sequences perceived as falling dipthongs are analysed as a vowel followed by /j/ or /w/ (see example~\ref{ex:falling-diphthongs}).

\begin{lingexample}
	\ex \label{ex:rising-diphthongs} /jɯl/ `year' \\
		/wætæn/ `homeland'
\end{lingexample}

\begin{lingexample}
	\ex \label{ex:falling-diphthongs} /ɑj/ `moon, month' \\
		/birɛw/ `someone, somebody'
\end{lingexample}

Length is not a contrastive properity, but it may arise from the monophthongisation of falling dipthongs involving high vowels.
This can happen in one of two ways, both of which require that the vowel be followed immediately by /j/ or /w/, and which result in the lengthening of the vowel.

The first case is where a high unrounded vowel is followed by /j/ or a high rounded vowel is followed by /w/.
This is accompanied by a drop of the consonant in question.

\begin{lingexample}
	\ex /gɾɔznɯj/ `Grozny' \becomes{} [gɾɔznɯː] \\
		/kij-/ `wear, put on' \becomes{} [kiː-] \\
		/suw/ `water' \becomes{} [suː]
\end{lingexample}

The second case is similar, but with the consonant in the following syllable\apostrophe{}s onset.
In this case, the consonant is not dropped.

\begin{lingexample}
	\ex /mɯjɯq/ `moustache' \becomes{} [mɯːjɯq] \\
		/suwum/ `my water' \becomes{} [suːwum].
\end{lingexample}

It is important to note that neither of these changes is reflected orthographically.

\section{Consonants}

Interturkic has 26 consonant phonemes, primarily classified by their voicing.
The unvoiced consonants are /p, t, k, q, ʔ, t͡ʃ, f, s, ʃ, x, h/.
The voiced consonants are /m, n, ŋ, b, d, g, d͡ʒ, v, z, ʒ, ɣ, w, l, j, ɾ/.

\begin{smalltable}{Consonant phonemes}{c|c|c|c|c|c}
	& \textbf{Labial} & \textbf{Alveolar} & \textbf{Postalveolar} & \textbf{Dorsal} & \textbf{Glottal} \\
	\hline
	\textbf{Nasal} & m & n & & ŋ & \\
	\hline
	\textbf{Stop} & p b & t d & & k q g & ʔ \\
	\hline
	\textbf{Affricate} & & & t͡ʃ d͡ʒ & & \\
	\hline
	\textbf{Fricative} & f v & s z & ʃ ʒ & x ɣ & h \\
	\hline
	\textbf{Approximant} & w & l & & j & \\
	\hline
	\textbf{Tap} & & ɾ & & & \\
\end{smalltable}

\chapter{Phonotactics}

\chapter{Morphophonological alternations}

\section{Final consonant lenition}

\section{Vowel harmony}

\section{Consonant assimilation}

\chapter{Stress}
