\chapter{Arabic}

\section{Main letters}

The Interturkic Arabic orthography has 34 main letters, as shown in table~\ref{%
tbl:arabic-alphabet}.
Owing to the script's cursive nature, a letter's shape changes based on its
position within a word.

Each letter has up to four positional forms, isolate (not joined on either
side), initial (joined to the following letter), medial (joined to both the
preceding and following letters), and final (joined to the preceding letter).
Letters that have only two positional forms have essentially fused their medial
forms into their final forms, and their initial forms into isolate forms.
Therefore, they do not connect to any following letter.

\begin{longtable}[c]{rllrrrr}
  \caption{Arabic alphabet}
  \label{tbl:arabic-alphabet}\\
  \toprule
  \multicolumn{2}{l}{Name}&\multirow[t]{2}{*}{IPA}
  &\multicolumn{4}{l}{Letter}\\
  \cmidrule(lr){1-2}\cmidrule(lr){4-7}
  \multicolumn{1}{l}{Arabic}&\multicolumn{1}{l}{Latin}&&\multicolumn{1}{l}{%
  final}&\multicolumn{1}{l}{medial}&\multicolumn{1}{l}{initial}&\multicolumn{1}
  {l}{isolate}\\
  \midrule
  \endfirsthead
  \arabtxt{الف}&\foreign{\latntxt{alıf}}&\phonem{\zero, ɑ}&\multicolumn{2}{r}{%
  \arabtxt{ـا}}&\multicolumn{2}{r}{\arabtxt{ا}}\\
  \arabtxt{بە}&\foreign{\latntxt{be}}&\phonem{b}&\arabtxt{ـب}&\arabtxt{ـبـ}&
  \arabtxt{بـ}&\arabtxt{ب}\\
  \arabtxt{پە}&\foreign{\latntxt{pe}}&\phonem{p}&\arabtxt{ـپ}&\arabtxt{ـپـ}&
  \arabtxt{پـ}&\arabtxt{پ}\\
  \arabtxt{تە}&\foreign{\latntxt{te}}&\phonem{t}&\arabtxt{ـت}&\arabtxt{ـتـ}&
  \arabtxt{تـ}&\arabtxt{ت}\\
  \arabtxt{ثە}&\foreign{\latntxt{se}}&\phonem{s}&\arabtxt{ـث}&\arabtxt{ـثـ}&
  \arabtxt{ثـ}&\arabtxt{ث}\\
  \arabtxt{جیم}&\foreign{\latntxt{cim}}&\phonem{d͡ʒ}&\arabtxt{ـج}&\arabtxt{ـجـ}&
  \arabtxt{جـ}&\arabtxt{ج}\\
  \arabtxt{چە}&\foreign{\latntxt{çe}}&\phonem{t͡ʃ}&\arabtxt{ـچ}&\arabtxt{ـچـ}&
  \arabtxt{چـ}&\arabtxt{چ}\\
  \arabtxt{اولو حە}&\foreign{\latntxt{ulu he}}&\phonem{h}&\arabtxt{ـح}&\arabtxt
  {ـحـ}&\arabtxt{حـ}&\arabtxt{ح}\\
  \arabtxt{خە}&\foreign{\latntxt{xe}}&\phonem{x}&\arabtxt{ـخ}&\arabtxt{ـخـ}&
  \arabtxt{خـ}&\arabtxt{خ}\\
  \arabtxt{دال}&\foreign{\latntxt{dal}}&\phonem{d}&\multicolumn{2}{r}{\arabtxt{%
  ـد}}&\multicolumn{2}{r}{\arabtxt{د}}\\
  \arabtxt{ذال}&\foreign{\latntxt{zal}}&\phonem{z}&\multicolumn{2}{r}{\arabtxt{%
  ـذ}}&\multicolumn{2}{r}{\arabtxt{ذ}}\\
  \arabtxt{رە}&\foreign{\latntxt{re}}&\phonem{ɾ}&\multicolumn{2}{r}{\arabtxt{ـر%
  }}&\multicolumn{2}{r}{\arabtxt{ر}}\\
  \arabtxt{زە}&\foreign{\latntxt{ze}}&\phonem{z}&\multicolumn{2}{r}{\arabtxt{ـز%
  }}&\multicolumn{2}{r}{\arabtxt{ز}}\\
  \arabtxt{ژە}&\foreign{\latntxt{je}}&\phonem{ʒ}&\multicolumn{2}{r}{\arabtxt{ـژ%
  }}&\multicolumn{2}{r}{\arabtxt{ژ}}\\
  \arabtxt{سین}&\foreign{\latntxt{sin}}&\phonem{s}&\arabtxt{ـس}
  &\arabtxt{ـسـ}&\arabtxt{سـ}&\arabtxt{س}\\
  \arabtxt{شین}&\foreign{\latntxt{şin}}&\phonem{ʃ}&\arabtxt{ـش}
  &\arabtxt{ـشـ}&\arabtxt{شـ}&\arabtxt{ش}\\
  \arabtxt{صاد}&\foreign{\latntxt{sad}}&\phonem{s}&\arabtxt{ـص}
  &\arabtxt{ـصـ}&\arabtxt{صـ}&\arabtxt{ص}\\
  \arabtxt{ضاد}&\foreign{\latntxt{zad}}&\phonem{z}&\arabtxt{ـض}
  &\arabtxt{ـضـ}&\arabtxt{ضـ}&\arabtxt{ض}\\
  \arabtxt{طا}&\foreign{\latntxt{ta}}&\phonem{t}&\arabtxt{ـط}
  &\arabtxt{ـطـ}&\arabtxt{طـ}&\arabtxt{ط}\\
  \arabtxt{ظا}&\foreign{\latntxt{za}}&\phonem{z}&\arabtxt{ـظ}
  &\arabtxt{ـظـ}&\arabtxt{ظـ}&\arabtxt{ظ}\\
  \arabtxt{عین}&\foreign{\latntxt{ayn}}&\phonem{\zero, ʔ}&\arabtxt{ـع}&\arabtxt
  {ـعـ}&\arabtxt{عـ}&\arabtxt{ع}\\
  \arabtxt{غین}&\foreign{\latntxt{ğayn}}&\phonem{ɣ}&\arabtxt{ـغ}&\arabtxt{ـغـ}&
  \arabtxt{غـ}&\arabtxt{غ}\\
  \arabtxt{فە}&\foreign{\latntxt{fe}}&\phonem{f}&\arabtxt{ـف}&\arabtxt{ـفـ}&
  \arabtxt{فـ}&\arabtxt{ف}\\
  \arabtxt{ڤە}&\foreign{\latntxt{ve}}&\phonem{v}&\arabtxt{ـڤ}&\arabtxt{ـڤـ}&
  \arabtxt{ڤـ}&\arabtxt{ڤ}\\
  \arabtxt{قاف}&\foreign{\latntxt{qaf}}&\phonem{q}&\arabtxt{ـق}&\arabtxt{ـقـ}&
  \arabtxt{قـ}&\arabtxt{ق}\\
  \arabtxt{کاف}&\foreign{\latntxt{kaf}}&\phonem{k}&\arabtxt{ـک}&\arabtxt{ـکـ}&
  \arabtxt{کـ}&\arabtxt{ک}\\
  \arabtxt{ساݣیر کاف}&\foreign{\latntxt{sañ\-ır kaf}}&\phonem{ŋ}&\arabtxt{ـݣ}&
  \arabtxt{ـݣـ}&\arabtxt{ݣـ}&\arabtxt{ݣ}\\
  \arabtxt{گاف}&\foreign{\latntxt{gaf}}&\phonem{g}&\arabtxt{ـگ}&\arabtxt{ـگـ}&
  \arabtxt{گـ}&\arabtxt{گ}\\
  \arabtxt{لام}&\foreign{\latntxt{lam}}&\phonem{l}&\arabtxt{ـل}&\arabtxt{ـلـ}&
  \arabtxt{لـ}&\arabtxt{ل}\\
  \arabtxt{میم}&\foreign{\latntxt{mim}}&\phonem{m}&\arabtxt{ـم}&\arabtxt{ـمـ}&
  \arabtxt{مـ}&\arabtxt{م}\\
  \arabtxt{نون}&\foreign{\latntxt{nun}}&\phonem{n}&\arabtxt{ـن}&\arabtxt{ـنـ}&
  \arabtxt{نـ}&\arabtxt{ن}\\
  \arabtxt{واو}&\foreign{\latntxt{waw}}&\phonem{w, u, y}&\multicolumn{2}{r}{%
  \arabtxt{ـو}}&\multicolumn{2}{r}{\arabtxt{و}}\\
  \arabtxt{کیچی هە}&\foreign{\latntxt{kiçi he}}&\phonem{h, ɑ, ɛ}&\arabtxt{ـه}&
  \arabtxt{ـهـ}&\arabtxt{هـ}&\arabtxt{ه}\\
  \arabtxt{یە}&\foreign{\latntxt{ye}}&\phonem{j, ɛ, ɯ, i}&\arabtxt{ـی}&\arabtxt
  {ـیـ}&\arabtxt{یـ}&\arabtxt{ی}\\
  \bottomrule
\end{longtable}

\foreign{\latntxt{Se}}, \foreign{\latntxt{ulu he}}, \foreign{\latntxt{zal}},
\foreign{\latntxt{sad}}, \foreign{\latntxt{zad}}, \foreign{\latntxt{ta}}, and
\foreign{\latntxt{za}} occur only in Perso\-/Arabic loans.
Word\-/initial \foreign{\latntxt{alıf}} and \foreign{\latntxt{ayn}} are silent.
When \foreign{\latntxt{kiçi he}} represents a vowel, it does not connect to the
following letter.

\section{Secondary letters}

Word\-/initial \phonem{ɑ} in non\-/Perso\-/Arabic items must be written with a
\arabtxt{ﻣﺪەلی الف} \foreign{\latntxt{meddeli alıf}}.
In Arabic loans, it may be used in non\-/initial position to represent \phonem{%
ʔɑ}.

\ex
  \vtop{\halign{
    \hfil\arabtxt{#}&\:\foreign{\latntxt{#}}\hfil&\:\transl{#}\hfil\cr
    آلتمیش&altmış&sixty\cr
    قرآن&Qur'an&(the) Qur'an\cr
  }}
\xe

In addition to \foreign{\latntxt{ayn}}, \arabtxt{ﻫﻤﺰە} \foreign{\latntxt{hemze}}
also represents \phonem{ʔ}.
Exclusively in word\-/final position of certain Arabic loans, it can be written
on its own~(\ref{ex:hemze-independent}).
It may also be written with \foreign{\latntxt{alıf}}, \foreign{\latntxt{waw}},
or \foreign{\latntxt{ye}}~(\ref{ex:hemze-bound}).
These graphemes are respectively named \arabtxt{همزەلی الف} \foreign{\latntxt{%
hemzeli alıf}}, \arabtxt{همزەلی واو} \foreign{\latntxt{hemzeli waw}}, and
\arabtxt{همزەلی یە} \foreign{\latntxt{hemzeli ye}}.

\pex
  \a\label{ex:hemze-independent}
    \vtop{\halign{
      \hfil\arabtxt{#}&\:\foreign{\latntxt{#}}\hfil&\:\transl{#}\hfil\cr
      ابتداء&ıbtıda'&beginning\cr
      جزء&cuz'&juz (one thirtieth of the Qur'an)\cr
    }}
  \a\label{ex:hemze-bound}
    \vtop{\halign{%
      \hfil\arabtxt{#}&\:\foreign{\latntxt{#}}\hfil&\:\transl{#}\hfil\cr
      مسألت&mas'alat&(a) problem\cr
      سؤال&su'al&question\cr
      فائدە&fa'ıda&gain, profit\cr
    }}
\xe

When suffixes are joined to words ending with an independent
\foreign{\latntxt{hemze}}, the \foreign{\latntxt{hemze}} must be turned into a
\foreign{\latntxt{hemzeli ye}}.

\foreign{\latntxt{Hemzeli alıf}} and \foreign{\latntxt{hemzeli waw}} are used
exclusively in Arabic loans.
\foreign{\latntxt{Hemzeli ye}} may be used in non\-/Arabic loans.
In fact, it is the correct way to represent \phonem{ʔ} in such items.

\ex
  \vtop{\halign{
    \hfil\arabtxt{#}&\:\foreign{\latntxt{#}}\hfil&\:\transl{#}\hfil\cr
    هاوایئی&Hawai'i&Hawaii (US state)\cr
    شیئان&Şi'an&Xi'an (city in China)\cr
  }}
\xe

\section{Ligatures}

When \foreign{\latntxt{lam}} is followed by \foreign{\latntxt{aıf}}, it must
be written with the ligature \foreign{\latntxt{lam\-/alıf}}~(\ref{ex:lam-alif}).
This is the only mandatory ligature in the Arabic orthography.

\ex\label{ex:lam-alif}
  \vtop{\halign{
    \hfil\arabtxt{#}&\:\foreign{\latntxt{#}}\hfil&\:\transl{#}\hfil\cr
    بولامن&bolaman&I become\cr
    هلال&hılal&crescent\cr
    آرسلان&arslan&lion\cr
  }}
\xe

When used to render the word for \transl{God} in monotheistic contexts, the
sequence \foreign{\latntxt{lam +~lam +~kiçi he}} must be written as \arabtxt{%
لله}~(\ref{ex:llah-with-alif}).
In compound terms involving the Arabic \arabtxt{لِـ} \foreign{li-} preposition,
no additional \foreign{\latntxt{lam}} is added~(\ref{ex:llah-without-alif}).

\pex
  \a\label{ex:llah-with-alif}
    \vtop{\halign{
      \hfil\arabtxt{#}&\:\foreign{\latntxt{#}}\hfil&\:\transl{#}\hfil\cr
      الله&Allah&God\cr
      بسم‌الله&bısmıllah&in the name of God\cr
      استغفرالله&astağfırullah&I seek forgiveness from God\cr
      ان‌شاءالله&ınşa'allah&if God wills it\cr
    }}
  \a\label{ex:llah-without-alif}
    \arabtxt{الحمد‌لله} \foreign{\latntxt{alhamdulıllah}} \transl{praise be to
    God}
\xe
