\chapter{Morphophonology}

Joining a suffix to a stem triggers a range of sound changes in both the stem
and suffix in question.
All of the changes described within this section are represented in writing.

\section{Last-vowel deletion}

Certain polysyllabic stems drop their last vowel if it is a high vowel, its
containing syllable is originally closed, and a suffix with an initial vowel is
added.

\ex
  \vtop{\halign{
    \phonem{#}\hfil&\:\transl{#}\hfil&\:\derivedto~\phonem{#}\hfil\cr
    ɑwuz&mouth&ɑwzum\cr
    buɾun&nose&buɾnum\cr
    fikiɾ&thought&fikɾim\cr
  }}
\xe

Whether or not this process applies to a particular stem is unpredictable.
each case must be memorised individually.

\section{Final-consonant lenition}

In native items, stem-final \phonem{p, k, q} become \phonem{b, g, ɣ}
respectively if a suffix with an initial vowel is added~(\ref{ex:lenition-yes}).
This does not happen if the final consonant in question is part of a complex
coda~(\ref{ex:lenition-no}).

\pex
  \a\label{ex:lenition-yes}
    \vtop{\halign{
      \phonem{#}\hfil&\:\transl{#}\hfil&\:+~\phonem{#}\hfil&\:\transl{(#)}\hfil%
      &\:\derivedto~\phonem{#}\hfil\cr
      tɑp-&find&\=/(I)l&\Pass&tɑbɯl-\cr
      syjɛk&bone&\=/(I)m&\Poss.\Fsg&syjɛgim\cr
      qulɑq&ear&\=/(s)I&\Poss.\Third&qulɑɣɯ\cr
    }}
  \a\label{ex:lenition-no}
    \vtop{\halign{
      \phonem{#}\hfil&\:\transl{#}\hfil&\:+~\phonem{#}\hfil&\:\transl{(#)}\hfil%
      &\:\derivedto~\phonem{#}\hfil\cr
      qɔɾq-&be afraid&\=/(I)t&\Caus&qɔɾqut-\cr
      tuɾp&radish&\=/(I)m&\Poss.\Fsg&tuɾpum\cr
    }}
\xe

\section{Vowel harmony}

Interturkic has a system of vowel harmony where the quality of a given vowel
determines the quality of the next vowel within the word.
This occurs both within stems and when suffixes are joined to a stem.

\subsection{Within stems}

With the exception of loans and compounds, words must consist entirely of front
vowels or of back vowels.
Furthermore, notwithstanding certain diphthongs (see~\ref{sec:diphthongs}),
rounded vowels must not follow unrounded vowels.

\subsection{Within suffixes}

Vowel harmony in Interturkic encompasses two mechanisms working in tandem:
backness harmony and rounding harmony.
Backness harmony requires a given vowel's backness to match that of the
preceding vowel.
Likewise, rounding harmony requires a given vowel's roundedness to match that of
the preceding vowel (note~\ref{sec:diphthongs}).
Derived from these two mechanisms are two types of vowel harmony, as in table~%
\ref{tbl:vowel-harmony}.

\begin{table}
  \centering
  \caption{Vowel harmony}
  \label{tbl:vowel-harmony}
  \begin{tabular}{lllll}
    \toprule
    \multirow[t]{2}{*}{Type}&\multicolumn{2}{l}{Front}
    &\multicolumn{2}{l}{Back}\\
    \cmidrule(lr){2-3}\cmidrule(lr){4-5}
    &unrounded&rounded&unrounded&rounded\\
    \midrule
    A&\multicolumn{2}{l}{\phonem{ɛ}}&\multicolumn{2}{l}{\phonem{ɑ}}\\
    I&\phonem{i}&\phonem{y}&\phonem{ɯ}&\phonem{u}\\
    \bottomrule
  \end{tabular}
\end{table}

A\=/type harmony alternates between \phonem{ɛ} in front\-/vowel environments
and \phonem{ɑ} in back\-/vowel environments, as in (\ref{ex:a-type-harmony})
with \phonem{\=/lAɾ} \transl{(\Pl)}.
I\=/type harmony alternates between \phonem{i} in front unrounded environments,
\phonem{y} in front rounded environments, \phonem{ɯ} in back unrounded
environments, and \phonem{u} in back rounded environments, as in (\ref{%
ex:i-type-harmony}) with \phonem{\=/nI} \transl{(\Acc)}.

\pex
  \a\label{ex:a-type-harmony}
    \vtop{\halign{
      \phonem{#}\hfil&\:\transl{#}\hfil&\:\derivedto~\phonem{#}\hfil\cr
      sœz&word&sœzlɛɾ\cr
      buzɑw&calf&buzɑwlɑɾ\cr
    }}
  \a\label{ex:i-type-harmony}
    \vtop{\halign{
      \phonem{#}\hfil&\:\transl{#}\hfil&\:\derivedto~\phonem{#}\hfil\cr
      ɛʃɛk&donkey&ɛʃɛkni\cr
      kymyʃ&silver&kymyʃny\cr
      ɑɣɑt͡ʃ&tree&ɑɣɑt͡ʃnɯ\cr
      jɔl&road, way&jɔlnu\cr
      biɾɛw&someone&biɾɛwny\cr
      tɑw&mountain&tɑwnu\cr
    }}
\xe

\section{Consonant assimilation}

Certain consonants in suffixes undergo assimilation in accordance with their
surrounding environments.
Two factors govern this process:
the voicing of the immediately preceding consonant and the backness of the
preceding vowel.
This leads to three possible types of consonant assimilation, as in table~\ref{%
tbl:consonant-assimilation}.

\begin{table}
  \centering
  \caption{Consonant assimilation}
  \label{tbl:consonant-assimilation}
  \begin{tabular}{lllll}
    \toprule
    \multirow[t]{2}{*}{Type}&\multicolumn{2}{l}{Unvoiced}
    &\multicolumn{2}{l}{Voiced}\\
    \cmidrule(lr){2-3}\cmidrule(lr){4-5}
    &unrounded&rounded&unrounded&rounded\\
    \midrule
    D&\multicolumn{2}{l}{\phonem{t}}&\multicolumn{2}{l}{\phonem{d}}\\
    G&\phonem{k}&\phonem{g}&\phonem{q}&\phonem{ɣ}\\
    K&\phonem{k}&\zero&\phonem{q}&\zero\\
    \bottomrule
  \end{tabular}
\end{table}

D\=/type assimilation depends on the first of the two factors, alternating
between \phonem{t} in unvoiced environments and \phonem{d} in voiced
environments, as in (\ref{ex:d-type-assimilation}) with \phonem{\=/DAn} \transl
{(\Abl)}.
K\=/type assimilation depends on the second factor, alternating between \phonem
{k} in front\-/vowel environments and \phonem{q} in back\-/vowel environments,
as in (\ref{ex:k-type-assimilation}) with \phonem{\=/lIK} \transl{(\Denom)}.
G\=/type assimilation depends on both factors, alternating between K\=/type
assimilation in unvoiced environments, \phonem{g} in front voiced environments,
and \phonem{ɣ} in back voiced environments, as in (\ref{ex:g-type-assimilation})
with \phonem{\=/GAn} \transl{(\Ptcp.\Pst)}.

\pex
  \a\label{ex:d-type-assimilation}
    \vtop{\halign{
      \phonem{#}\hfil&\:\transl{#}\hfil&\quad \derivedto~\phonem{#}\hfil\cr
      bulut&cloud&buluttɑn\cr
      biz&we&bizdɛn\cr
    }}
  \a\label{ex:k-type-assimilation}
    \vtop{\halign{
      \phonem{#}\hfil&\:\transl{#}\hfil&\:\derivedto~\phonem{#}\hfil&\:\transl{%
      #}\hfil\cr
      biɾ&one&biɾlik&unity\cr
      dɔst&friend&dɔstluq&friendship\cr
    }}
  \a\label{ex:g-type-assimilation}
    \vtop{\halign{
      \phonem{#}\hfil&\:\transl{#}\hfil&\:\derivedto~\phonem{#}\hfil\cr
      kɛt-&go&kɛtkɛn\cr
      jɑt-&lie&jɑtqɑn\cr
      kœɾ-&see&kœɾgɛn\cr
      bɔl-&become&bɔlɣɑn\cr
    }}
\xe

If there is no immediately preceding consonant, then the environment is
considered voiced for the purposes of this process.
