\chapter{Verbs}

\section{Stems}

Every verb has two stems: affirmative and negative.
The affirmative stem is formed by removing the infinitive suffix from a verb's
citation form, i.e.\ being identical to its bare stem~(\vref{ex:aff-stem}).
The negative stem is formed by suffixing \foreign{\latntxt{\morphophon{\=/mA}}}
onto an affirmative stem~(\vref{ex:neg-stem}).

\pex
  \a\label{ex:aff-stem}%
    \vtop{\halign{
      \foreign{\latntxt{#}}\hfil&\:\transl{to #}\hfil&\:\derivedas~\foreign
      {\latntxt{#\=/}}\hfil\cr
      kelmek&come&kel\cr
      yazmaq&write&yaz\cr
    }}
  \a\label{ex:neg-stem}%
    \vtop{\halign{
      \foreign{\latntxt{#-}}\hfil&\:\transl{#}\hfil&\:\derivedas~\foreign
      {\latntxt{#\=/}}\hfil&\:\transl{not #}\cr
      kel&come&kel\=/me&come\cr
      yaz&write&yaz\=/ma&write\cr
    }}
\xe

All finite and non\-/finite forms are built on top of one of these two stems.

\section{Grammatical categories}

\subsection{Tense}

Tense expresses \emph{where} on the timeline an event is situated.

The past, present, and future are absolute tenses, referring respectively to
events before, at, and after the time of speech.
The distant past is a relative tense, expressing a time far before the time of
speech, often relative to another event that itself is in the past.

\subsection{Aspect}

Aspect expresses \emph{how} an event extends over the timeline, orthogonal to
\emph{when} it occurs.
Interturkic envisages an event as having two limits, one at its beginning, and
another at its culmination or end.
The exact nature of these limits is depends on a particular verb's semantics.

Intraterminal items refer to events between these two limits, comparable to
imperfectives in other languages.
Postterminal items refer to events after their attainment of some critical
point, comparable to perfectives in other languages.
Terminal items refer to events precisely at one of these two limits.

\subsection{Focality}

Focality expresses the degree to which the speaker perceives the event as being
of current relevance.
It is often\---but not always\---conditioned by aspect.

High\-/focal items refer to events contained within a window of time narrow
enough to be considered of current and direct relevance.
Low\-/focal items refer to events whose occurences span over a period of time
wide enough to not be considered immediately relevant.

\section{Non\-/finite forms}

\subsection{Infinitives}

Infinitives are formed with the suffix \foreign{\latntxt{\morphophon{\=/mAK}}}.
They are used when the verb is mentioned as a lexical item and in certain
constructions.

\ex
\xe

\subsection{Converbs}

Converbs are verb forms that take on adverbial functions, particularly in
subordination to other clauses.

Intraterminal converbs are formed with the suffix \foreign{\latntxt{\morphophon
{\=/A}}} on consonant\-/final stems and \foreign{\latntxt{\morphophon{\=/y}}} on
vowel\-/final stems.
They denote events\---both one\-/off and repeated\---that occur simultaneously
with that of the superordinate verb.

\ex
  \begingl
    \gla Ya'qub bölme\=/ge kül\=/e kir\=/di\=/\zero.//
    \glb Ya'qub room\=/\Dat{} laugh\=/\Cvb.\Intra{} enter\=/\Pst\=/\Tsg//
    \glft\transl{Ya'qub entered the room laughing} (he was laughing as he came
    in)//
  \endgl
\xe

Postterminal converbs are formed with the suffix \foreign{\latntxt{\morphophon
{\=/(I)p}}}.
They denote events that start/occur, one after the other, with the last in the
sequence represented by the main (finite) verb.

\ex
  \begingl
    \gla Ol üy\=/ge kel\=/ip, qol\=/lar\=/ı\=/n yuw\=/ma\=/p, yaralan\=/dı%
    \=/\zero.//
    \glb \Tsg{} home\=/\Dat{} come\=/\Cvb.\Post{} hand\=/\Pl\=/\Poss.\Third%
    \=/\Acc{} wash\=/\Neg{}\=/\Cvb.\Post{} wound\_self\=/\Pst\=/\Tsg//
    \glft \transl{He came home, did not wash his hands, and wounded himself.}//
  \endgl
\xe

\subsection{Participles}

Participles are verb forms that take on adjectival functions.
They may either qualify noun phrases, or themselves become nominalised and
accept nominal inflections.

\subsubsection{Past participles}

Past participles are formed with the suffix \foreign{\latntxt{\morphophon
{\=/GAn}}}.
They denote events that took place in the past, relative to the time of speech,
whether or not the results thereof are relevant or still in effect.
Especially with stative verbs and verbs of motion, present\=/tense relevance may
be implied.

\ex
  \begingl
    \gla Yan\=/ıñ\=/da otur\=/ğan tabıb kim?//
    \glb side\=/\Poss.\Ssg\=/\Loc{} sit\=/\Ptcp.\Pst{} doctor who//
    \glft \transl{Who is that doctor sitting (lit.\ who has sat) next to you?}//
  \endgl
\xe

\subsubsection{Present participles}

Low\-/focal present participles are formed with the suffixes \foreign{\latntxt{%
\morphophon{\=/A\-dur\-ğan}}} and \foreign{\latntxt{\morphophon{\=/y\-dur%
\-ğan}}}
(\derivedfrom~\foreign{\latntxt{\morphophon{\=/A/y} tur\=/ğan}} \inlinegls{\Cvb%
.\Intra{} stand\=/\Ptcp.\Pst}), the former on consonant\-/final stems and the
latter on vowel\-/final stems.
They refer to events that, at the time of speech, take place habitually or as a
general matter of fact, often with the implication they will continue in the
future.

\ex
  \begingl
    \gla Her kün çetreng oyna\=/ydurğan bala\=/lar üy\=/ge ket\=/me\=/di%
    \=/\zero.//
    \glb every day chess play\=/\Ptcp.\Prs.\Lfoc{} children\=/\Pl{} home\=/\Dat
    {} go\=/\Neg\=/\Pst\=/\Tpl//
    \glft \transl{The children who play chess every day have not gone home
    (yet).}//
  \endgl
\xe

High\-/focal present participles are formed with the suffixes \foreign{\latntxt
{\morphophon{\=/Ayatqan}}} and \foreign{\latntxt{\morphophon{\=/yyatqan}}}
(\derivedfrom~\foreign{\latntxt{\morphophon{\=/A/y} yat\=/qan}} \inlinegls{\Cvb%
.\Intra{} lie\=/\Ptcp.\Pst}), the former on consonant\-/final stems and the
latter on vowel\-/final stems.
They refer to events currently in progress at the time of speech.

\ex
  \begingl
    \gla Seni üyret\=/eyatqan qatın Ortaqtürk til\=/i\=n bil\=/e\=/di.//
    \glb \Ssg.\Acc{} teach\=/\Ptcp.\Prs.\Hfoc{} woman Interturkic language%
    \=/\Tsg\=/\Acc{} know\=/\Prs.\Lfoc\=/\Tsg//
    \glft \transl{The woman teaching you (right now) knows Interturkic.}//
  \endgl
\xe

\subsubsection{Future participles}

Low\-/focal future participles are formed with the suffix \foreign{\latntxt{%
\morphophon{\=/(A)r}}} in the affirmative and \foreign{\latntxt{\=/z}} in the
negative.
They express durative properties with a general belief that they will continue
into the future.

\ex
  \begingl
    \gla Yan\=/ar taw\=/nuñ iç\=/i\=/nde köp sön\=/me\=/z ot bar.//
    \glb burn\=/\Ptcp.\Fut.\Lfoc{} mountain\=/\Gen{} inside\=/\Poss.\Tsg\=/\Loc
    {} many go\_out\=/\Neg\=/\Ptcp.\Fut.\Lfoc{} fire exist//
    \glft \transl{There are many inextinguishable (lit.\ not\-/going\-/out)
    fires inside the volcano (lit.\ burning mountain).}//
  \endgl
\xe

High\-/focal future participles are formed with the suffix \foreign{\latntxt{%
\morphophon{\=/(A)cAK}}}.
They express events about which the speaker is certain will happen in the
future.

\ex
  \begingl
    \gla Bügün kel\=/ecek kitab\=/lar her talıb üçün\=/\zero.//
    \glb today come\=/\Ptcp.\Fut.\Hfoc{} book\=/\Pl{} every student for\=/\Cop%
    .\Prs.\Tpl//
    \glft \transl{The books that will come today are for every student.}//
  \endgl
\xe

\section{Finite forms}
